00-Data-Processing-Example

### 2020-11-17 examples/00-Data-Processing-Example.Rmd
v1 Added routing process from Chris' accessibility notebooks

@@

Load packages used in the notebook:
```{r message = FALSE}
library(pycno)
library(readr)
library(rgrass7)
library(sf)
library(tidyverse)
```

@@

@@

Adjust bounding box to remove docking station in park, which is for tourism purposes, to the east:
```{r}
bounding_box[3] <- 597200
```

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

## Provenance of data

@@

Load packages used in the notebook:
```{r message = FALSE}
library(disk.frame)
library(pycno)
library(readr)
library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
library(sf)
library(tidyverse)
```

Define some parameters for `disk.frame` and `r5r`:
```{r}
setup_disk.frame()
options(scipen = 999)
options(java.parameters = "-Xmx6G")
options(future.globals.maxSize = Inf)
```

### Read data

@@

Plot service areas:
```{r}
ggplot() +
  geom_sf(data = sobi_service,
          aes(fill = AREA_NAME))
```

There is a Core Service Area and a much smaller service area at Van Wagner's. This is a location at the beach with service for beach goers. We will remove this area from further analysis:
```{r}
sobi_service <- sobi_service %>%
  dplyr::filter(AREA_NAME == "Core Service Area")
```

@@

@@

### Extract features in service area

@@

### Read population

@@

### Visualize geographical data

@@

### Prepare geographical data for interpolation

@@

### Remove all features that are not population from the DAs

@@

## Interpolate population

@@

## Extract population in service area

@@

Note that some cells have population values of zero:
```{r}
interpolated_pop_clean %>% select(population) %>% summary()
```

Filter all cells with non-zero population values:
```{r}
interpolated_pop_clean <- interpolated_pop_clean %>%
            dplyr::filter(population > 0)
```

Add identifier for the population cells:
```{r}
interpolated_pop_clean <- interpolated_pop_clean %>%
            mutate(UID = 1:n())
```

@@

## Bikeshare system

Read Sobi hubs:
```{r}
sobi_hubs <- st_read("SoBi_Hubs.shp")
```

Plot interpolated population and SoBi hubs:
```{r}
ggplot() + 
  geom_sf(data = interpolated_pop_clean,
          aes(color = population)) +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = sobi_hubs) +
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1)
```

Remove SoBi hub on beach:
```{r}
sobi_hubs <- sobi_hubs %>%
  dplyr::filter(NAME != "Van Wagners")
```

Plot again:
```{r}
ggplot() + 
  geom_sf(data = interpolated_pop_clean,
          aes(color = population)) +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = sobi_hubs,
          aes(size = RACKS_AMOU)) +
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1) +
  scale_size(range = c(1, 4))
```

## Routing

I used [BBBike](https://download.bbbike.org/osm/bbbike/) to extract OSM data for Hamilton - not the whole city, but enough to cover SoBI's service area. The name of the file is `planet_-80.042,43.183_-79.748,43.309.osm.pbf`. Copy to folder `r5_graph`.

Set Up R5 Routing. First define the path to where the graph is located:
```{r set up r5 path, include=FALSE}
r5_path <- file.path("./r5_graph")
```

Build the graph:
```{r build-graph, include = FALSE}
r5_sobi <- setup_r5(data_path = r5_path, verbose = FALSE)
```

Prepare Input Data for `r5r`. The origins are the coordinates of the population cells and the destinations the coordinates of the SoBi hubs:
```{r prepare-inputs}
# save origin centroids in format expected by R5R (id, lon, lat)
origins_i <- data.frame(UID = interpolated_pop_clean$UID, 
              interpolated_pop_clean %>%
                st_transform(crs = 4326) %>%
                st_coordinates()) %>%
    rename(lon = X, lat = Y, id = UID) %>%
    select(id, lon, lat)

# now SoBi hubs
destinations_j <- data.frame(OBJECTID = sobi_hubs$OBJECTID, 
              sobi_hubs %>%
                st_transform(crs = 4326) %>%
                st_coordinates()) %>%
    rename(lon = X, lat = Y, id = OBJECTID) %>%
    select(id, lon, lat)
```

Calculate OD Matrix for Walking:
```{r calculate walking od matrix, include=FALSE}
# set up batching according to how many origin rows to process at one time
chunksize = 2000 # larger chunks for walking will give enough origins in each chunk to allow multiprocessing to spin up with R5R
num_chunks = ceiling(nrow(origins_i)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(origins_i,
                          outdir = "./df/origins_i",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  origins_i_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5_sobi,
                          origins = origins_i_chunk,
                          destinations = destinations_j,
                          mode = c("WALK"),
                          departure_datetime = as.POSIXct(strptime("2020-03-02 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_walk_dist = 10000, # metres
                          max_trip_duration = 30)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "./df/output_ttm_walk",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
ttm_walk_time <- end.time - start.time
ttm_walk_time
```

## Extract travel time matrix

```{r load od matrix for walk and subway, include=FALSE}
# connect to the walking travel time matrix disk frame
ttm_walk.disk.frame <- disk.frame("./df/output_ttm_walk")
#opportunities_j <- as.data.frame(nyc_cb_point) %>% select(GEOID10, total_emp) %>% rename(toId = GEOID10, o_j = total_emp)
```

Convert disk.frame to data frame:
```{r}
ttm_walk <- as.data.frame(ttm_walk.disk.frame) %>%
  transmute(UID = as.numeric(fromId), OBJECTID = as.numeric(toId), travel_time)
```

## Save data to disk

Rename data frame:
```{r}
population_sobi <- interpolated_pop_clean
```

Save data files:
```{r}
save(population_sobi, file = "population_sobi.RData")
save(sobi_hubs, file = "sobi_hubs.RData")
save(sobi_service, file = "sobi_service.RData")
save(ttm_walk, file = "ttm_walk.RData")
```
